---
title: "COVID-19 Cases vs. Mortality"
output: html_notebook
---
I have a macabre fascination with tracking the course of the COVID-19 pandemic. I suspect there are two reasons for this. One, by delving into the numbers I imagine I have some control over this thing.  Second, it feels like lighting a candle to show that science can reveal truth at a time when the darkness of anti-science is creeping across the land. 

The purpose of this project is, as usual, twofold.  First, to explore an interesting data science question and, second, to explore some techniques and packages in the R universe.  We will be looking at the relationship of COVID-19 cases to mortality.  What is the lag between a positive case and a death?  How does that vary among states?  How has it varied as the pandemic has progressed?

I have been thinking about how to measure mortality lags for a while now.  What prompted to do a write-up was discovering a new function in Matt Dancho's `timetk` package, `tk_augment_lags`, which makes short work of building multiple lags.  Not too long ago, managing models for multiple lags and multiple states would have been a bit messy.  The emerging "tidy models" framework using "list columns" is immensely powerful for this sort of thing.  I was almost giddy ("Nerd!") when I was able to reduce so much analysis into so few lines of code.

## Introduction

There is no shortage of data sets to work with. Here we will use the NY Times COVID tracking data set which is updated daily.  The package `covid19nytimes` lets us refresh the data on demand.


```{r Initialize}
# correlate deaths and cases by state
library(tidyverse)
library(covid19nytimes)
library(timetk)
library(lubridate)
library(broom)

# source https://github.com/nytimes/covid-19-data.git
us_states_long <- covid19nytimes::refresh_covid19nytimes_states()
save(us_states_long,file="data/us_states_long.rdata")
us_states_long %>% glimpse()
```

The NY Times data is presented in a "long" format.  When we start modeling, long will suit us well but first we have to add features to help us and that will require `pivot'ing to wide, adding features and then back to long.  The daily data is so irregular the first features we will add are 7-day moving averages to smooth the series.  We'll also do a nation-level analysis first so we aggregate the state data as well.

```{r clean data, message=FALSE, warning=FALSE}
# Create rolling average changes
# pivot wider
# this will also be needed when we create lags
us_states <- us_states_long %>%
  # discard dates before cases were tracked.
  filter(date > as.Date("2020-03-01")) %>% 
  pivot_wider(names_from="data_type",values_from="value") %>% 
  rename(state=location) %>%
  select(date,state,cases_total,deaths_total) %>%
  mutate(state = as_factor(state)) %>% 
  arrange(state,date) %>% 
  group_by(state) %>%
  #smooth the data with 7 day moving average
  mutate(cases_7day = (cases_total - lag(cases_total,7))/7) %>%
  mutate(deaths_7day = (deaths_total - lag(deaths_total,7))/7) %>%
  {.}

# national analysis
# ----------------------------------------------
# aggregate state to national
us <- us_states %>%
  group_by(date) %>% 
  summarize(across(.cols=where(is.double),
                   .fns = function(x)sum(x,na.rm = T),
                   .names="{.col}"))

us %>% glimpse()
```
Let's look at the cumulative case and death counts in the U.S.  It is bad form to put lines with different scales on a plot but it's worthwhile to compare the slopes of each line.  New cases (in yellow) are accelerating but mortality (in red) is not. 
```{r show cumulative, message=TRUE, warning=FALSE}
coeff <- 30
us %>% 
  ggplot(aes(date,cases_total)) + geom_line(color="orange") +
  theme(legend.position = "none") +
  geom_line(aes(y=deaths_total*coeff),color="red") +
  scale_y_continuous(labels = scales::comma,
                     name = "Cases",
                     sec.axis = sec_axis(deaths_total~./coeff,
                                         name="Deaths",
                                         labels = scales::comma)) +
  theme(
    axis.title.y = element_text(color = "orange", size=13),
    axis.title.y.right = element_text(color = "red", size=13)
  ) +
  labs(title =  "U.S. Cases vs. Deaths",
       x = "Date")
```
We might be tempted to simply regress deaths vs. cases but a scatter plot shows us that would not be satisfactory.  "Satisfactory" is a word I will use a lot in this post.  There is so much we are still learning about this virus I won't be able to draw highly definitive conclusions.  "Satisfactory" will have to be good enough.

```{r}
# does a simple scatterplot tell us anything 
# about the relationship of deaths to cases? No.
us %>% 
  ggplot(aes(deaths_7day,cases_7day)) + geom_point()
```
We can get much more insight plotting smoothed deaths and cases over time.  A couple of observations are obvious.  First when cases start to rise, deaths follow with a lag.  Second, we have had three spikes in cases so far and in each successive instance the mortality has risen by a smaller amount.  This suggests that, thankfully, we are getting better at treating this disease. It is NOT a function of increased testing because [positivity rates](http://91-divoc.com/pages/covid-visualization/?chart=countries&highlight=United%20States&show=highlight-only&y=highlight&scale=linear&data=testPositivity-daily-7&data-source=merged&xaxis=right-12wk#countries) have not been falling.


```{r}
#visualize the relationship between rolling average of weekly scases and deaths
us %>% 
  ggplot(aes(date,cases_7day)) + geom_line(color="orange") +
  theme(legend.position = "none") +
  geom_line(aes(x=date,y=deaths_7day*coeff),color="red") +
  scale_y_continuous(labels = scales::comma,
                     name = "Cases",
                     sec.axis = sec_axis(deaths_7day~./coeff,
                                         name="Deaths",
                                         labels = scales::comma)) +
  theme(
    axis.title.y = element_text(color = "orange", size=13),
    axis.title.y.right = element_text(color = "red", size=13)
  ) +
  labs(title =  "U.S. Cases vs. Deaths",
       subtitle = "7-Day Average",
       x = "Date")
```
This illustrates a problem for any modeling we might do.It looks like the more cases surge the less the impact on deaths.  This is NOT a valid conclusion. The passage of time has more explanatory power than cases in predicting deaths so we have to take that into account.

```{r}
# passage of time affects deaths more than cases
lm(deaths_7day~cases_7day+date,data=us) %>% summary()
```


```{r message=FALSE, warning=FALSE}
#create columns for deaths led 0 to 30 days ahead
us_lags <- us %>%
  # create lags by day
  tk_augment_lags(deaths_7day,.lags = -60:0,.names="auto") %>% 
  # create lags by week is using weekly data
  # tk_augment_lags(deaths_7day,.lags = -8:0,.names="auto") %>% 
  {.}
# fix names to remove minus sign
names(us_lags) <- names(us_lags) %>% str_replace_all("lag-","lead")
us_lags[1:10] %>% glimpse()
```


nest-map-unnest workflow

```{r}
# make long form to nest
# initialize models data frame
models <- us_lags %>% ungroup %>% 
  pivot_longer(cols = contains("lead"),
               names_to = "lead",
               values_to = "led_deaths") %>% 
  select(date,cases_7day,lead,led_deaths) %>% 
  mutate(lead = as.numeric(str_remove(lead,"deaths_7day_lead"))) %>% 
  {.}

# make separate tibbles for each regression
models <- models %>% 
  nest(data=c(date,cases_7day,led_deaths)) %>% 
  arrange(lead)

#Run a linear regression on lagged cases and date vs deaths

# Polynomial degree for date.
degree = 1

models <- models %>% 
  mutate(model = map(data,
                     function(df) 
                       lm(led_deaths~cases_7day+poly(date,degree),data = df)))


# Add regression coefficient
# get adjusted r squared
models <- models %>% 
  mutate(adj_r = map(model,function(x) glance(x) %>% 
                       pull(adj.r.squared))
         %>% unlist)

# Show model fit by lead time
# make predictions using best model
best_fit <- models %>% 
  summarize(adj_r = max(adj_r)) %>% 
  left_join(models)

models %>%
  ggplot(aes(lead,adj_r)) + geom_line() +
  labs(subtitle = paste("Best fit lead =",best_fit$lead,"days"),
       title = "Model Fit By Lag Days",
       x = "Lead Time in Days for Deaths",
       y= "Adjusted R-squared")
```


```{r}
# ------------------------------------------
# see how well our model predicts

# Function to create prediction plot
show_predictions <- function(single_model){
  complete_cases <- single_model$data[[1]] %>% 
    remove_missing()
  
  predictions <- enframe(predict(single_model$model[[1]],na.action = exclude),
                         name = NULL,
                         value = paste0("predicted_in_","ahead","_days"))
  
  bind_cols(complete_cases,predictions) %>% 
    rename(cases=cases_7day,
           deaths_actual=led_deaths,
           deaths_predicted=predicted_in_ahead_days) %>%
    select(-cases) %>% 
    pivot_longer(cols = where(is.numeric)) %>% 
    ggplot(aes(date,value,color=name)) + geom_line()
  
}

# best adj_r is not the best fit
show_predictions(best_fit)
```

Override best fit
```{r}
# local maximum of 14 days gives  much more satisfactory result
lag_override <- 14
show_predictions(models[lag_override,])
```


```{r}
# ------------------------------------------
# state by state analysis
# states hit their peaks (to date) at very different times, making state
# a useful feature. We could use one dummy variable for each state but it
# is simpler with tidymodels to conduct a separate set of models for each state

# illustrate selected states
us_states %>% 
  filter(state %in% c("Florida","Texas","California","Michigan")) %>% 
  ggplot(aes(date,cases_7day)) + geom_line(color="orange") +
  facet_wrap(~state,scales = "free") +
  theme(legend.position = "none") +
  geom_line(aes(y=deaths_7day*coeff),color="red") +
  scale_y_continuous(labels = scales::comma,
                     name = "Cases",
                     sec.axis = sec_axis(deaths_7day~./coeff,
                                         name="Deaths",
                                         labels = scales::comma)) +
  theme(
    axis.title.y = element_text(color = "orange", size=13),
    axis.title.y.right = element_text(color = "red", size=13)
  ) +
  labs(title =  "U.S. Cases vs. Deaths",
       subtitle = "7-Day Average",
       x = "Date")
```


```{r message=FALSE, warning=FALSE}
# create lags
us_states_lags <- us_states %>%
  # create lags by day
  tk_augment_lags(deaths_7day,.lags = -60:0,.names="auto") %>% 
  # create lags by week is using weekly data
  # tk_augment_lags(deaths_7day,.lags = -8:0,.names="auto") %>% 
  {.}
# fix names to remove minus sign
names(us_states_lags) <- names(us_states_lags) %>% str_replace_all("lag-","lead")

# make long form to nest
# initialize models data frame
models_st <- us_states_lags %>% ungroup %>% 
  pivot_longer(cols = contains("lead"),
               names_to = "lead",
               values_to = "led_deaths") %>% 
  select(state,date,cases_7day,lead,led_deaths) %>% 
  mutate(lead = as.numeric(str_remove(lead,"deaths_7day_lead"))) %>% 
  {.}

# make separate tibbles for each regression
models_st <- models_st %>% 
  nest(data=c(date,cases_7day,led_deaths)) %>% 
  arrange(lead)

#Run a linear regression on lagged cases and date vs deaths
# Polynomial degree for date.
degree = 1

models_st <- models_st %>% 
  mutate(model = map(data,
                     function(df) 
                       lm(led_deaths~cases_7day+poly(date,degree),data = df)))


# Add regression coefficient
# get adjusted r squared
models_st <- models_st %>% 
  mutate(adj_r = map(model,function(x) glance(x) %>% 
                       pull(adj.r.squared))
         %>% unlist)

models_st %>%
  ggplot(aes(lead,adj_r)) + geom_line() +
  facet_wrap(~state)
```


```{r}
# best fit lag by state
best_fit <- models_st %>% 
  group_by(state) %>% 
  summarize(adj_r = max(adj_r)) %>% 
  left_join(models_st)

best_fit %>% ggplot(aes(adj_r)) + 
  geom_histogram(bins = 10,color="white") +
  geom_vline(xintercept = models$adj_r[lag_override],color="red") +
  annotate(geom="text",x=0.5,y=12,label="Adj-R in National Model") +
  labs(y = "State Count",
       x="Adjusted R-Squared",
       title = "Goodness of Fit of State Models",
       caption = "Source:NY Times,My Analysis")
```


```{r}
best_fit %>% ggplot(aes(lead)) + 
  geom_histogram(binwidth = 5,color="white") +
  labs(y = "State Count",
    x="Best Fit Model Days from Case to Death",
    title = "COVID-19 Lag Time From Cases to Death",
    caption = "Source:NY Times,My Analysis")
```


```{r}
# ----------------------------------------------------
# Reality check with longitudinal data from Ohio

models_st %>% 
  filter(state=="Ohio") %>% 
  ggplot(aes(lead,adj_r)) + geom_line() +
  labs(x = "Lead Time in Days for Deaths",
       y= "Adjusted R-squared",
       title = "Ohio: Model Fit by Lag Days")
```


```{r}
# source: https://coronavirus.ohio.gov/static/COVIDSummaryData.csv
ohio_raw <- read_csv("https://coronavirus.ohio.gov/static/COVIDSummaryData.csv", 
                     col_types = cols(`Admission Date` = col_date(format = "%m/%d/%Y"), 
                                      `Date Of Death` = col_date(format = "%m/%d/%Y"), 
                                      `Onset Date` = col_date(format = "%m/%d/%Y")))


fix_df_colnames <- function(df){
  names(df)<-names(df) %>% 
    str_replace_all(c(" " = "_" , "," = "" )) %>% 
    tolower()
  return(df)
}

ohio <- ohio_raw %>% 
  fix_df_colnames() %>%
  mutate(sex = as_factor(sex),
         county = as_factor(county),
         age_range = as_factor(age_range))


ohio <- ohio %>% mutate(onset_to_death = as.numeric(date_of_death - onset_date),
                        fatal_flag = if_else(!is.na(date_of_death),TRUE,FALSE),
                        onset_year = year(onset_date),
                        onset_week = epiweek(onset_date))
  

onset_to_death <- ohio %>%
  filter(county != "Grand Total") %>% 
  group_by(onset_year,onset_week) %>%
  summarise(mean_onset_to_death = mean(onset_to_death,na.rm = TRUE)) %>%
  mutate(date=as.Date(paste(onset_year,onset_week,1),"%Y %U %u")) %>% 
  remove_missing()

# helper function to annotate plots 
pos_index <- function(index_vec,fraction){
  return(index_vec[round(length(index_vec)*fraction)])
}
avg_lag <- mean(onset_to_death$mean_onset_to_death)
onset_to_death %>% ggplot(aes(date,mean_onset_to_death)) + 
  geom_col() +
  geom_hline(yintercept = avg_lag) +
  annotate(geom="text",
           label=paste("Average Lag =",round(avg_lag)),
           y=20,x=pos_index(onset_to_death$date,.8)) +
  labs(x = "Onset Date",
       y = "Mean Onset to Death",
       title = "Ohio Days from Illness Onset Until Death Over Time",
       subtitle = paste("Average =",
                        round(mean(onset_to_death$mean_onset_to_death)),"Days"))

```

